A general and efficient representation of Ancestral Recombination Graphs

### 2022-08-26 paper.tex
Cleanups on history section

Recent breakthroughs in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,wohns2022unified}
and data representation~\citep{kelleher2016efficient}
have raised the realistic prospect of genealogical analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using these inferred ancestries as input have begun to
appear~\citep{osmond2021estimating,zhang2021biobank,fan2022genealogical}
and many more are sure to
follow~\citep{harris2019database}. This vibrant research area, however,
has a significant difficulty, which risks hampering
progress: we currently lack a well-defined data-model and shared terminology
to discuss recombining genealogies. This leads to basic errors in statements
about what different inference methods produce as output, as well as presenting serious
difficulties in either comparing the outputs of different methods or
(from a user's perspective) using multiple inference methods in an analysis.

@@

In the 1990s, Griffiths and colleagues revisited the
coalescent with recombination from a different perspective,
in which the stochastic process itself is encoded as a graph
(with particular properties; see section XXX for details),
and showed how the local trees generated by Hudson's algorithm
were embedded within
it~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
They referred to this graph,
representing both the stochastic process and
its random realisations, as the Ancestral Recombination Graph (ARG).
Although mathematically equivalent, it is
important to note that the Griffiths and Hudson formulations of
the coalescent with recombination are not identical;
in particular, a direct implementation of the ARG process
requires exponential time to simulate
(see Appendix XXX for details). However, ARGs provided a way
to describe recombinant ancestry as single object that
could be inferred in a way
that was missing from Hudson's treatments.
An overview of the Hudson and Griffiths stochastic processes is provided in the appendix.

Early work on ARG inference focused on the problem of
inferring parameters of the
stochastic process, where the ancestry was regarded as a
latent parameter to be averaged out
\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
fearnhead2001estimating}.
These methods met with limited success
because the state space of ARGs is overwhelmingly large and
lacks a simple geometry or neighbourhood structure for inference or
sampling methods to  exploit.
The focus subsequently shifted to the more
tractable---but still NP-hard~\citep{wang2001perfect}---problem of computing
the minimum number of recombinations required
to explain an input dataset~\citep{myers2003bounds}, and inferring the corresponding
ARG realisations~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
It is important to note that such recombination-parsimony ARGs
have quite different properties from realisations of the
ARG \emph{process} as modelled by Hudson, Griffiths, and colleagues.
Firstly, branch lengths are not typically estimated,
% why do we need the following two lines
and node times are required in order to compute a likelihood under the
coalescent with recombination (see section XXX).
Secondly, even if branch lengths were estimated,
the realisations would have a very small likelihood since
the coalescent with recombination is often highly \emph{un}parsimonious
in terms of recombination events for non-trivial scaled recombination rate.
The previously close correspondance
between the ARG as a mathematical representation of the coalescent process
and as concrete realisations of ancestry was now therefore broken.
\cite{minichiello2006mapping} explicitly distinguished the ARG as a
stochastic process from the ARG as a data structure, and
argued for using the term to mean specific realised ancestral histories.
This has become a common
usage~\citep{mathieson2020ancestry,brandt2021evaluation}
[TODO expand sentence].
Gusfield~\citep{gusfield2014recombinatorics}
defines also defines an ARG as a concrete realisation
and studies instances that satisfy particular
properties such as ``galled trees''
\citep{wang2001perfect, gusfield2004optimal}
as well as describing algorithms for more general phylogenetic
networks~\citep{huson2010phylogenetic}.

% JK: This paragraph is weak, and needs some work. Basically want to
% illustrate the mathematical literature which views ancestral
% graphs very much as *processes*.
% Jere: ASG section redrafted, gene transfer left as TODO (because I
% know essentially nothing about it).
In parallel with this tendency to decouple the term
Ancestral Recombination Graph from its original defining stochastic process,
an extensive mathematical literature has developed in
studying other related stochastic processes encoded via a graph.
In particular, the Ancestral Selection Graph
(ASG)~\citep{krone1997ancestral,neuhauser1997genealogy}
uses a similar approach to model natural selection.
Unlike the ARG process, the ASG imposes a hard distinction between the stochastic process,
which constructs a random ARG-like graph, and an observable realisation,
which is a single tree sampled from the graph in a non-uniform way to encode desired
patterns of natural selection.
Constructions of ASG-like stochastic processes encoding various
forms of selection, often in parallel with recombination or other genetic forces,
are an area of considerable and ongoing theoretical interest~\citep[e.g.][]{
neuhauser1999ancestral,
donnelly1999genealogical,
fearnhead2001perfect,
fearnhead2003ancestral,
etheridge2009coalescent,
gonzalezcasanova2018duality,
koskela2019robust}.
Similarly, the Ancestral Gene Transfer Graph

models horizontal gene transfer~\citep{baumdicker2014infinitely}
by [FINISH ME.

The point of this historical overview is to illustrate that the term
Ancestral Recombination Graph has different meanings in different
contexts, and in its original form it was closely coupled to
the coalescent with recombination. In the following sections we
show that, although the ARG-as-a-data-structure perspective is
now the most common interpretation, the \emph{form} of that
data structure is still closely derived from the original
stochastic process, and that this form is inflexible
in the types of genetic inheritance that can be described,
computationally inefficient, and
forces complete (and potentially spurious) precision about
the location and timing
of recombination events. % (and coalescent events?)
Unless otherwise stated, we will use the ARG-as-a-data-structure
interpretation for the remainder of this paper.

>>>>

Recent breakthroughs in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,wohns2022unified}
and data representation~\citep{kelleher2016efficient}
have raised the realistic prospect of genealogical analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using these inferred ancestries as input have begun to
appear~\citep{osmond2021estimating,zhang2021biobank,fan2022genealogical,hejase2022deep}
and many more are sure to
follow~\citep{harris2019database}. This vibrant research area, however,
has a significant difficulty, which risks hampering
progress: we currently lack a well-defined data-model and shared terminology
to discuss recombining genealogies. This leads to basic errors in statements
about what different inference methods produce as output, as well as presenting serious
difficulties in either comparing the outputs of different methods or
(from a user's perspective) using multiple inference methods in an analysis.

@@

In the 1990s, Griffiths and colleagues revisited the
coalescent with recombination from a different perspective,
in which the stochastic process itself is encoded as a graph,
and showed how the local trees generated by Hudson's algorithm
were embedded within
it~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
They referred to this graph,
representing both the stochastic process and
its random realisations, as the Ancestral Recombination Graph (ARG).
Although mathematically equivalent, it is
important to note that the Griffiths and Hudson formulations of
the coalescent with recombination are not identical;
in particular, a direct implementation of the ARG process
requires exponential time to simulate
(see Appendix XXX for details). However, ARGs provided a way
to describe recombinant ancestry as single object that
could be reasoned about and inferred in a way
that was missing from Hudson's treatments.

Early work on ARG inference focused on the problem of
inferring parameters of the
stochastic process, where the ancestry was regarded as a
latent parameter to be averaged out
\citep[e.g.][]{griffiths1996ancestral,kuhner2000maximum, nielsen2000estimation,
fearnhead2001estimating}.
These methods met with limited success
because the state space of ARGs is overwhelmingly large and
lacks a simple geometry or neighbourhood structure for inference or
sampling methods to  exploit.
The focus subsequently shifted to the more
tractable---but still NP-hard~\citep{wang2001perfect}---problem of computing
the minimum number of recombinations required
to explain an input dataset~\citep{myers2003bounds}, and inferring the corresponding
ARG realisations~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
It is important to note that such recombination-parsimony ARGs
have quite different properties from realisations of the
ARG \emph{process} as modelled by Hudson, Griffiths, and colleagues.
Firstly, branch lengths are not typically estimated,
% why do we need the following two lines
and node times are required in order to compute a likelihood under the
coalescent with recombination (see section XXX).
Secondly, even if branch lengths were estimated,
the realisations would have a very small likelihood since
the coalescent with recombination is often highly \emph{un}parsimonious
in terms of recombination events.
The previously close correspondance
between the ARG as a mathematical representation of the coalescent process
and as concrete realisations of ancestry was therefore broken.
\cite{minichiello2006mapping} explicitly distinguished the ARG as a
stochastic process from the ARG as a data structure, and
argued for using the term to mean specific realised ancestral histories.
This has become a common
usage~\citep[e.g.][]{gusfield2014recombinatorics,mathieson2020ancestry,brandt2021evaluation},
although not universal, with authors also using the term
in the original stochastic
process sense~\citep[e.g.][]{nordborg2000linkage,birkner2013ancestral,
wilton2015smc,griffiths2016coalescent}.

% Gusfield~\citep{gusfield2014recombinatorics}
% defines also defines an ARG as a concrete realisation
% and studies instances that satisfy particular
% properties such as ``galled trees''
% \citep{wang2001perfect, gusfield2004optimal}
% as well as describing algorithms for more general phylogenetic
% networks~\citep{huson2010phylogenetic}.

% JK: This paragraph is weak, and needs some work. Basically want to
% illustrate the mathematical literature which views ancestral
% graphs very much as *processes*.
% Jere: ASG section redrafted, gene transfer left as TODO (because I
% know essentially nothing about it).
In parallel with this tendency to decouple the term
Ancestral Recombination Graph from the stochastic process,
an extensive mathematical literature has developed in
studying other related stochastic processes encoded via a graph.
In particular, the Ancestral Selection Graph
(ASG)~\citep{krone1997ancestral,neuhauser1997genealogy}
uses a similar approach to model natural selection.
Unlike the ARG process, the ASG imposes a hard distinction between the stochastic process,
which constructs a random ARG-like graph, and an observable realisation,
which is a single tree sampled from the graph in a non-uniform way to encode desired
patterns of natural selection.
Constructions of ASG-like stochastic processes encoding various
forms of selection, often in parallel with recombination or other genetic forces,
are an area of considerable and ongoing theoretical interest~\citep[e.g.][]{
neuhauser1999ancestral,
donnelly1999genealogical,
fearnhead2001perfect,
fearnhead2003ancestral,
etheridge2009coalescent,
gonzalezcasanova2018duality,
koskela2019robust}.

The point of this historical overview is to illustrate that the term
Ancestral Recombination Graph has different meanings in different
contexts, and in its original form it was closely coupled to
the coalescent with recombination. In the following sections we
show that, although the ARG-as-a-data-structure perspective is
now the most common interpretation, the \emph{form} of that
data structure is still closely linked to the original
stochastic process, and that this form is inflexible
in the types of genetic inheritance that can be described,
computationally inefficient, and
forces complete (and potentially spurious) precision about
the location and timing
of recombination events. % (and coalescent events?)
Unless otherwise stated, we will use the ARG-as-a-data-structure
interpretation for the remainder of this paper.
